Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-03 linear_adia.tex
updating

\subsection{Perturbations with thermal relaxation in the thin-disk limit}\label{analytic_relax}
We now introduce thermal relaxation ($\beta>0$). We pursue explicit
solutions in the thin-disk approximation. The governing equation,
Eq. \ref{vertiso_gov_nondim}, is then 
\begin{align}
  0 = \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    \epsilon q\hat{k}\right), 
\end{align}
and $\chi =
\left(1-\ii\hat{\sigma}\beta\right)/\left(1-\ii\hat{\sigma}\beta\gamma\right)
$ in terms of dimensionless variables.

@@

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about a known 
solution for $\beta\equiv0$  ($\chi\equiv 1$). 
Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta.
\end{align}
Here $\delta v_z$ corresponds to the explicit solution developed in
\S\ref{iso_poly} with  $\hat{\sigma}^2$ given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}  and keeping
only first order terms, we obtain 
\begin{align}
 0 =& \delta v_{z1}^{\prime\prime} - z\left(1 + \ii\epsilon q
   k\right)\delta v_{z1}^\prime\notag\\
 &+ \left[2\hat{\sigma}\delta\hat{\sigma}\left(1 + \khat^2\right) +
   \left(\hat{\sigma}^2 - 1\right)\delta\chi \right.\notag\\
&\phantom{++} \left. + \zhat^2\left(\khat^2 -
     \ii\epsilon q \khat\right)\delta\chi\right]\delta v_z\notag\\
 &+\left[\hat{\sigma}^2\left(1 + \khat^2\right) - \left(1+\ii\epsilon
     q \khat\right)\right]\delta v_{z1}.\label{nearly_iso4}
\end{align}

@@

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}. These
studies reached this conclusion by considering vertically-localized
disturbances. % , although a definitive timescale was not presented. 
By contrast, our Eq. \ref{iso_vsi_cond} quantifies this requirement
for vertically global disturbances (i.e. the fundamental mode). We see
that in a thin disk, a very short thermal timescale
 is required ($\beta_\mathrm{crit}\ll 1$) because
$\beta_\mathrm{crit}\propto \epsilon$. 
This suggests that
finite thermal relaxation rapidly stabilizes the VSI.

>>>>

\subsection{Perturbations with thermal relaxation in the thin-disk limit}\label{analytic_relax}
We now introduce thermal relaxation ($\beta>0$), but pursue explicit
solutions in the thin-disk approximation for simplicity. The governing
equation, Eq. \ref{vertiso_gov_nondim}, is then 
\begin{align}
  0 = \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    \epsilon q\hat{k}\right), 
\end{align}
and $\chi =
\left(1-\ii\hat{\sigma}\beta\right)/\left(1-\ii\hat{\sigma}\beta\gamma\right)
$ in terms of dimensionless variables.

@@

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about a known 
solution for $\beta\equiv0$  ($\chi\equiv 1$). 
Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta.
\end{align}
Here $\delta v_z$ corresponds to an explicit solution developed in
\S\ref{iso_poly} with  $\hat{\sigma}^2$ given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}  and keeping
only first order terms, we obtain 
\begin{align}
 0 =& \delta v_{z1}^{\prime\prime} - z\left(1 + \ii\epsilon q
   k\right)\delta v_{z1}^\prime\notag\\
 &+ \left[2\hat{\sigma}\delta\hat{\sigma}\left(1 + \khat^2\right) +
   \left(\hat{\sigma}^2 - 1\right)\delta\chi \right.\notag\\
&\phantom{++} \left. + \zhat^2\left(\khat^2 -
     \ii\epsilon q \khat\right)\delta\chi\right]\delta v_z\notag\\
 &+\left[\hat{\sigma}^2\left(1 + \khat^2\right) - \left(1+\ii\epsilon
     q \khat\right)\right]\delta v_{z1}.\label{nearly_iso4}
\end{align}

@@

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}, but
this conclusion was reached in these studies by considering
vertically-localized disturbances. % , although a definitive timescale
                                % was not presented.  
By contrast, our Eq. \ref{iso_vsi_cond} quantifies this requirement
for vertically global disturbances (i.e. the fundamental mode). We see
that in a thin disk, a very short thermal timescale
 is required ($\beta_\mathrm{crit}\ll 1$) because
$\beta_\mathrm{crit}\propto \epsilon$. 
This suggests that
finite thermal relaxation rapidly stabilizes the VSI.

### 2015-03-03 linear_iso.tex
updating

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. We  first examine isothermal perturbations ($\beta =
0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.

\subsection{Relation to the vertically global
  shearing box} 
We connect the linearized equations 
Eq. \ref{lin_mass}---\ref{lin_energy} to the vertically global  
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to
allow the shear flow in the shearing box to be height-dependent.  
The difference between the radially-local approximation and the VGSB is
that the radial disk structure is ignored in the latter. That is, the 
equilibrium density and pressure field in the VGSB is homogeneous in
the radial ($x$) direction, as in the standard shearing box.

Thus, the linear problem in the VGSB can be derived from
Eq. \ref{lin_mass}---\ref{lin_energy} by setting 
$\hat{g}_c=0$. The resulting linear problem is very similar 
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply some of their solution methods.

@@

\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  % &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
  %     \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
  &+ \left[\frac{\sigma^2}{c_s^2}\left(\frac{k_x^2c_s^2}{\Omega_k^2} +
      \chi\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
  &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
    \left(1-\chi\right) 
  \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. We examine isothermal perturbations ($\beta =
0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.

\subsection{Relation to the vertically global
  shearing box} 
We connect the linearized equations 
Eq. \ref{lin_mass}---\ref{lin_energy} to the vertically global  
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to
allow the shear flow in the shearing box to be height-dependent. This
is an attempt to capture the effect of vertical shear in a local model
of accretoin disks. The difference between the radially-local
approximation and the VGSB is that the radial disk structure is
ignored in the latter. That is, the  equilibrium density and pressure
field in the VGSB is homogeneous in the radial ($x$) direction, as in
the standard shearing box.

Thus, the linear problem in the VGSB can be derived from
Eq. \ref{lin_mass}---\ref{lin_energy} by setting 
$\hat{g}_c=0$. The resulting problem is very similar 
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply some of their solution methods.

@@

\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ for $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  % &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
  %     \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
  &+ \left[\frac{\sigma^2}{c_s^2}\left(\frac{k_x^2c_s^2}{\Omega_k^2} +
      \chi\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
  &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
    \left(1-\chi\right) 
  \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

### 2015-03-03 paper.tex
updating

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the
  disk  midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in a stably stratified disk. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include an energy
  equation with a thermal relaxation term that restores the disk
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebulae (MMSN) with an 
  estimate for the physical thermal timescales 
  in the MMSN based on dust opacity. We find the MMSN is unstable to
  the VSI at tens of AU with a characteristic growth times of $\sim
  30$ local orbital periods.  
  % This suggests the VSI is dynamically important in
  % the outer parts of protoplanetary disks.  
\end{abstract}

@@

However, MHD turbulence may not be applicable for all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. In this respect, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

%recent work on vsi 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple interpretation for the VSI: an
exchange of fluid parcels while conserving their angular momenta
individually, may result in a new configuration of lower energy, if
the exchange occurs over a sufficiently large vertical displacement.

@@

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
temperature profile of the disk essentially remains unchanged during
the disk evolution. This has been confirmed by 
\cite{stoll14}, who find the VSI is only sustained in radiation
hydrodynamic simulations with the inclusion of heating from stellar
irradiation, which resulted in an effective thermal timescale $t_c\ll
P_\mathrm{orb}$.

@@

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually assumed for typical disk
models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk be unstable to the VSI, it must pass through the
linear instability first. Thus, understanding the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03}, which accounts for thermal
relaxation, but considered vertically localized disturbances. Our
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations which, in fact, we find to be the
dominant mode in protoplanetary disks with a realistic estimate for
the thermodynamic timescale.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include a thermal
  relaxation term that restores the disk  
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI at tens of AU with
  a characteristic growth times of $\sim 30$ local orbital periods. 
\end{abstract}

@@

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. In this respect, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

%recent work on vsi 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This dependence
occurs when the equilibrium pressure depends on both the temperature
and density, making the disk baroclinic. It is thus generally
expected that $\p_z\Omega\neq 0$ in astrophysical disks.

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple interpretation for the VSI: an
exchange of fluid parcels while conserving their angular momenta
individually, may result in a new configuration of lower energy, if
the exchange occurs over a sufficiently large vertical displacement.

@@

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
disk temperature essentially needs to remain unchanged during its
evolution. This was further supported with radiation hydrodynamic
simulations by \cite{stoll14}, who find the VSI is only sustained with
the inclusion of heating from stellar irradiation, which resulted in
an effective thermal timescale $t_c\ll P_\mathrm{orb}$.

@@

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually
assumed for typical disk models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk become turbulent through the VSI, it must pass through the
linear instability first. Thus, understanding the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of earlier 
analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
evolution, but considered vertically localized disturbances. Our 
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations which, in fact, we find to be the
dominant mode in protoplanetary disks with a realistic estimate for
the thermodynamic timescales.
